decompr: GVC decomposition in R

### 2015-01-26 README.Rmd
recompile

---
title: "decompr: trade flow decomposition for Global Value Chain analysis"
author: "Bastiaan Quast and Victor Kummritz"
date: "Wednesday, November 26, 2014"
output: md_document
bibliography: bibliography.bib
---

@@

Abstract
===============
The R package **decompr** implements algorithms for the analysis of Global Value Chains.
Two methods for gross export flow decomposition using Inter-Country Input-Output tables are provided.
The first method is the **Wang-Wei-Zhu** [-@wang2014quantifying] algorithm which splits bilateral gross exports into 16 value added components.
These components can broadly be divided into domestic and foreign value added.
The second method concerns a **source decomposition** algorithm.
This derives the value added origin of a country-industry???s exports by source country and source industry,
by applying the basic Leontief insight to gross trade data.
This article summarises the methodology of the algorithms and describes the format of the input and output data.

>>>>

---
title: "decompr: trade flow decomposition for Global Value Chain analysis"
author: "Bastiaan Quast and Victor Kummritz"
date: "Wednesday, November 26, 2014"
bibliography: bibliography.bib
---

@@

Abstract
===============
The R package **decompr** implements algorithms for the analysis of Global Value Chains.
Two methods for gross export flow decomposition using Inter-Country Input-Output tables are provided.
The first method is the **Wang-Wei-Zhu** [-@wang2014quantifying] algorithm which splits bilateral gross exports into 16 value added components.
These components can broadly be divided into domestic and foreign value added.[@wang2014quantifying]
The second method concerns a **source decomposition** algorithm.
This derives the value added origin of a country-industry???s exports by source country and source industry,
by applying the basic Leontief insight to gross trade data.
This article summarises the methodology of the algorithms and describes the format of the input and output data.

### 2015-01-26 draft.Rnw
recompile

\begin{document}
\maketitle

@@

\section{Data}
Two data sets are included in the package, namely
the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world} and, 
furthermore, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify
the mechanics of the package.

@@

\subsection{Output}
<<lt-out, out.extra='angle=90' >>=
lt
@

>>>>

\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle

@@

\section{Data}
Two data sets are included in the package, one real world data set, and one minimal fictional data, for training surposes.
The forme is the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
The later, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify
the mechanics of the package.

@@

\subsection{Output}
<<lt-out >>=
lt
@

### 2015-01-26 draft.tex
recompile

\begin{document}
\maketitle

\vspace{\stretch{1}}
\abstract{The R package \textit{decompr} implements algorithms for the analysis of Global Value Chains.
Two methods for gross export flow decomposition using Inter-Country Input-Output tables are provided.
The first method is the \textit{Wang-Wei-Zhu} algorithm which splits bilateral gross exports into 16 value added components.
These components can broadly be divided into domestic and foreign value added.
The second method concerns a decomposition based on the classical \textit{Leontief} insight.
This derives the value added origin of a country-industrys exports by source country and source industry,
by applying the basic Leontief insight to gross trade data.
This article summarises the methodology of the algorithms and describes the format of the input and output data.}
\pagebreak

\section{Introduction}
Global Value Chains (GVCs) have become a central factor in trade and development policy.
Policy makers from different countries and institutions have placed them at the center of their agenda
and continuously emphasize their growing importance for both international trade and economic development.
However, analysing this phenomenon empirically requires complex matrix manipulations,
since the relevant data is only available in the form of gross flows.
This package enables researchers with little background in matrix algebra and linear programming
to easily derive standard GVC indicators for statistical analysis.

As mentioned above, \textit{decompr} uses Inter-Country Input-Output tables,
such as those published by the OECD, WIOD, or national statistics bureaus.
These tables state input-output relationships in gross terms
between industries within and across countries.

@@

\section{Data}
Two data sets are included in the package.
The WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
Furthermore, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify the mechanics of the package.

<<load-package, message=FALSE >>=
# load the package
library(decompr)
@

<<load-data >>=
# load the data
data(leather)
# list the objects in the data set
ls()
@

@@

\begin{align}
\begin{split}
V (I-A)^{-1}E=&
\begin{pmatrix}
v_{i}^{k}& 0& 0& 0\\
0& v_{j}^{k}& 0& 0\\
0& 0& v_{i}^{l}& 0\\
0& 0& 0& v_{j}^{l}
\end{pmatrix}
*
\begin{pmatrix}
b_{ii}^{kk}& b_{ij}^{kk}& b_{ii}^{kl}& b_{ij}^{kl}\\
b_{ji}^{kk}& b_{jj}^{kk}& b_{ji}^{kl}& b_{jj}^{kl}\\
b_{ii}^{lk}& b_{ij}^{lk}& b_{ii}^{ll}& b_{ij}^{ll}\\
b_{ji}^{lk}& b_{jj}^{lk}& b_{ji}^{ll}& b_{jj}^{ll}
\end{pmatrix}
*
\begin{pmatrix}
e_{i}^{k}& 0& 0& 0\\
0& e_{j}^{k}& 0& 0\\
0& 0& e_{i}^{l}& 0\\
0& 0& 0& e_{j}^{l}
\end{pmatrix}
 \\
=& 
\begin{pmatrix}
v_{i}^{k}b_{ii}^{kk}e_{i}^{k}& v_{i}^{k} b_{ij}^{kk}e_{j}^{k}& v_{i}^{k}b_{ii}^{kl}e_{i}^{l}& v_{i}^{k}b_{ij}^{kl}e_{j}^{l}\\
v_{j}^{k}b_{ji}^{kk}e_{i}^{k}& v_{j}^{k}b_{jj}^{kk}e_{j}^{k}& v_{j}^{k}b_{ji}^{kl}e_{i}^{l}& v_{j}^{k}b_{jj}^{kl}e_{j}^{l}\\
v_{i}^{l}b_{ii}^{lk}e_{i}^{k}& v_{i}^{l}b_{ij}^{lk}e_{j}^{k}& v_{i}^{l}b_{ii}^{ll}e_{i}^{l}& v_{i}^{l}b_{ij}^{ll}e_{j}^{l}\\
v_{j}^{l}b_{ji}^{lk}e_{i}^{k}& v_{j}^{l}b_{jj}^{lk}e_{j}^{k}& v_{j}^{l}b_{ji}^{ll}e_{i}^{l}& v_{j}^{l}b_{jj}^{ll}e_{j}^{l}
\end{pmatrix}
=
\begin{pmatrix}
vax_{ii}^{kk}& vax_{ij}^{kk}& vax_{ii}^{kl}& vax_{ij}^{kl}\\
vax_{ji}^{kk}& vax_{jj}^{kk}& vax_{ji}^{kl}& vax_{jj}^{kl}\\
vax_{ii}^{lk}& vax_{ij}^{lk}& vax_{ii}^{ll}& vax_{ij}^{ll}\\
vax_{ji}^{lk}& vax_{jj}^{lk}& vax_{ji}^{ll}& vax_{jj}^{ll}
\end{pmatrix}
\end{split}
\end{align}
with
\begin{align*}
&v_{s}^{c} = \frac{va_{s}^{c}}{y_{s}^{c}} = 1 - \sum_{d}^{2}{a_{kc}^{ds}} - \sum_{d}^{2}{a_{lc}^{ds}} \hspace{2cm} (c \in k,l \hspace{.5cm} s \in i,k),\\
&\begin{pmatrix}
b_{ii}^{kk}& b_{ij}^{kk}& b_{ii}^{kl}& b_{ij}^{kl}\\
b_{ji}^{kk}& b_{jj}^{kk}& b_{ji}^{kl}& b_{jj}^{kl}\\
b_{ii}^{lk}& b_{ij}^{lk}& b_{ii}^{ll}& b_{ij}^{ll}\\
b_{ji}^{lk}& b_{jj}^{lk}& b_{ji}^{ll}& b_{jj}^{ll}
\end{pmatrix}
=
\begin{pmatrix}
1-a_{ii}^{kk}& -a_{ij}^{kk}& -a_{ii}^{kl}& -a_{ij}^{kl}\\
-a_{ji}^{kk}& 1-a_{jj}^{kk}& -a_{ji}^{kl}& -a_{jj}^{kl}\\
-a_{ii}^{lk}& -a_{ij}^{lk}& 1-a_{ii}^{ll}& -a_{ij}^{ll}\\
-a_{ji}^{lk}& -a_{jj}^{lk}& -a_{ji}^{ll}& 1-a_{jj}^{ll}
\end{pmatrix}^{-1},
\end{align*}
and
\begin{equation*}
a_{cf}^{su} = \frac{inp_{cf}^{su}}{y_{f}^{u}}  \hspace{2cm} (c,f \in k,l \hspace{.5cm} s,u \in i,k),
\end{equation*}
where $v_{s}^{c}$ gives the share of value added, $va_{s}^{c}$, in output , $y_{s}^{c}$, and $e_{s}^{c}$ indicates gross exports. Finally, $a_{cf}^{su}$ denotes the share of inputs , $inp_{cf}^{su}$, in output. The elements of the $V(I-A)^{-1}E$ or $vax$ matrix are our estimates for the country-industry level value added origins of each country-industry's exports. decompr implements this algorithm into R to automate the process of deriving the matrix. Equipped with it, researchers can calculate standard GVC indicators. Examples include \citet{dahuetal01}'s Vertical Specialisation ratio at the industry-level which can be derived by summing for each country and industry across the value added of all foreign countries and industries. Alternatively, the four dimensions of the matrix (source country, source industry, using country, using industry) allow for industry-level gravity-type estimations of value added trade flows.

@@

<<load_tables >>=
decompr_object <- load_tables_vectors( x = inter,
                                       y = final,
                                       k = countries,
                                       i = industries,
                                       o = out         )
@

We can now decompose the elements using the Leontief decomposition.
<<leontief >>=
lt <- leontief( decompr_object )
@

@@

<<leontief-wrapper >>=
lt2 <- decomp( x = inter,
               y = final,
               k = countries,
               i = industries,
               o = out,
               method = "leontief" )
@

@@

\subsection{Output}
<<lt-out, out.extra='angle=90' >>=
lt
@

@@

<<wwz >>=
w <- wwz(decompr_object)
@

>>>>

\usepackage{Sweave}
\begin{document}
\input{draft-concordance}
\maketitle

\vspace{\stretch{1}}
\abstract{Global Value Chains have become a central unit of analysis in research on international trade. However, 
the complex matrix transformations at the basis of most Value Chain indicators still constitute a significant entry barrier
the field. The R package \textit{decompr} solves this problem by implementing algorithms for the analysis of Global Value Chains,
which calculate standard indicators in an automated fashion. Two methods for gross export flow decomposition using 
Inter-Country Input-Output tables are provided. The first method is the \textit{Wang-Wei-Zhu} algorithm which splits 
bilateral gross exports into 16 value added components. These components can broadly be divided into domestic and 
foreign value added in exports. The second method concerns a decomposition based on the classical \citet{wale36} insight.
It derives the value added origins of an industry's exports by source country and source industry, using easily available
gross trade data. This article summarises the methodology of the algorithms, describes the format of the input and output data, 
and exemplifies the usefulness of the two methods on the basis of a simple example data set.}
\pagebreak

\section{Introduction}
The analysis of Global Value Chains (GVCs) has become a central topic in research on trade and development policy.
Both policy makers and academia increasingly value the growth opportunities GVCs offer to world trade and, especially, 
to developing countries. However, analysing this phenomenon empirically requires complex matrix manipulations,
since the relevant data is only available in the form of gross flows.
This package enables researchers with little background in matrix algebra and linear programming
to easily derive standard GVC indicators for statistical analysis.

\textit{decompr} uses Inter-Country Input-Output tables (ICIOs), such as those published by the OECD, 
the World Input Output Database (WIOD),  or national statistics bureaus, as input. 
These tables state supply and demand relationships in gross terms between industries within and across countries.

@@

\section{Data}
Two data sets are included in the package, one real world data set, and one minimal fictional data, for training surposes.
The forme is the WIOD regional Inter-Country Input Output tables for the year 2011 \parencite{timmer2012world}. 
The later, a fictional 3-country 3-sector data set is included, which we will use throughout this article to exemplify
the mechanics of the package.

\begin{Schunk}
\begin{Sinput}
> # load the package
> library(decompr)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
> # load the data
> data(leather)
> # list the objects in the data set
> ls()
\end{Sinput}
\begin{Soutput}
[1] "countries"  "final"      "industries" "inter"      "out"       
\end{Soutput}
\end{Schunk}

The data is set up in order to illustrate the benefits of the decompositions. That is, we follow the flows of intermediate
goods through a fictional GVC and show how the readily available gross trade flows differ from the decomposed value added
flows. To this end, we construct the elements of the input-output tables such that we have two countries and two industries
that focus on upstream tasks, which means they focus on supplying other industries, and one industry and one country that
is specialized in downstream tasks, that is it serves mainly final demand. In our example the upstream industries are 
Agriculture and Textiles while the downstream industry in Transport Equipment. Similarly, Argentina and Turkey represent 
upstream countries with Germany being located downstream within this specific value chain.

@@

\begin{align}
\begin{split}
V (I- A &)^{-1} E=
\begin{pmatrix}
v_{k}^{i}& 0& 0& 0\\
0& v_{k}^{j}& 0& 0\\
0& 0& v_{l}^{i}& 0\\
0& 0& 0& v_{l}^{j}
\end{pmatrix}
*
\begin{pmatrix}
b_{kk}^{ii}& b_{kk}^{ij}& b_{kl}^{ii}& b_{kl}^{ij}\\
b_{kk}^{ji}& b_{kk}^{jj}& b_{kl}^{ji}& b_{kl}^{jj}\\
b_{lk}^{ii}& b_{lk}^{ij}& b_{ll}^{ii}& b_{ll}^{ij}\\
b_{lk}^{ji}& b_{lk}^{jj}& b_{ll}^{ji}& b_{ll}^{jj}
\end{pmatrix}
*
\begin{pmatrix}
e_{k}^{i}& 0& 0& 0\\
0& e_{k}^{j}& 0& 0\\
0& 0& e_{l}^{i}& 0\\
0& 0& 0& e_{l}^{j}
\end{pmatrix}
=
\\
&
\begin{pmatrix}
v_{k}^{i}b_{kk}^{ii}e_{k}^{i}& v_{k}^{i} b_{kk}^{ij}e_{k}^{j}& v_{k}^{i}b_{kl}^{ii}e_{l}^{i}& v_{k}^{i}b_{kl}^{ij}e_{l}^{j}\\
v_{k}^{j}b_{kk}^{ji}e_{k}^{i}& v_{k}^{j}b_{kk}^{jj}e_{k}^{j}& v_{k}^{j}b_{kl}^{ji}e_{l}^{i}& v_{k}^{j}b_{kl}^{jj}e_{l}^{j}\\
v_{l}^{i}b_{lk}^{ii}e_{k}^{i}& v_{l}^{i}b_{lk}^{ij}e_{k}^{j}& v_{l}^{i}b_{ll}^{ii}e_{l}^{i}& v_{l}^{i}b_{ll}^{ij}e_{l}^{j}\\
v_{l}^{j}b_{lk}^{ji}e_{k}^{i}& v_{l}^{j}b_{lk}^{jj}e_{k}^{j}& v_{l}^{j}b_{ll}^{ji}e_{l}^{i}& v_{l}^{j}b_{ll}^{jj}e_{l}^{j}
\end{pmatrix}
=
\begin{pmatrix}
vae_{kk}^{ii}& vae_{kk}^{ij}& vae_{kl}^{ii}& vae_{kl}^{ij}\\
vae_{kk}^{ji}& vae_{kk}^{jj}& vae_{kl}^{ji}& vae_{kl}^{jj}\\
vae_{lk}^{ii}& vae_{lk}^{ij}& vae_{ll}^{ii}& vae_{ll}^{ij}\\
vae_{lk}^{ji}& vae_{lk}^{jj}& vae_{ll}^{ji}& vae_{ll}^{jj}
\end{pmatrix}
\end{split}
\end{align}
\begin{align*}
&v_{c}^{s} = \frac{va_{c}^{s}}{y_{c}^{s}} = 1 - a_{kc}^{is} - a_{kc}^{js} - a_{lc}^{js} - a_{lc}^{is} \hspace{2cm} (c \in k,l \hspace{.5cm} s \in i,j),\\
&\begin{pmatrix}
b_{kk}^{ii}& b_{kk}^{ij}& b_{kl}^{ii}& b_{kl}^{ij}\\
b_{kk}^{ji}& b_{kk}^{jj}& b_{kl}^{ji}& b_{kl}^{jj}\\
b_{lk}^{ii}& b_{lk}^{ij}& b_{ll}^{ii}& b_{ll}^{ij}\\
b_{lk}^{ji}& b_{lk}^{jj}& b_{ll}^{ji}& b_{ll}^{jj}
\end{pmatrix}
=
\begin{pmatrix}
1-a_{kk}^{ii}& -a_{kk}^{ij}& -a_{kl}^{ii}& -a_{kl}^{ij}\\
-a_{kk}^{ji}& 1-a_{kk}^{jj}& -a_{kl}^{ji}& -a_{kl}^{jj}\\
-a_{lk}^{ii}& -a_{lk}^{ij}& 1-a_{ll}^{ii}& -a_{ll}^{ij}\\
-a_{lk}^{ji}& -a_{lk}^{jj}& -a_{ll}^{ji}& 1-a_{ll}^{jj}
\end{pmatrix}^{-1},
\end{align*}
and
\begin{equation*}
a_{cf}^{su} = \frac{inp_{cf}^{su}}{y_{f}^{u}}  \hspace{2cm} (c,f \in k,l \hspace{.5cm} s,u \in i,j),
\end{equation*}
where \(v_{s}^{c}\) gives the share of industry \emph{s}'s value added,
\(va_{c}^{s},\) in output, \(y_{s}^{c}\), and \(e_{k}^{i}\)
indicates gross exports. Finally, \(a_{su}^{cf}\) denotes
the share of inputs, \(inp_{su}^{cf}\), in output. 
The elements of the $V(I-A)^{-1}E$ or $vae$ matrix are our estimates for the country-industry level value added
origins of each country-industry's exports. \textit{decompr} implements this algorithm into R to automate the process of
deriving the matrix. Equipped with it, researchers can calculate standard GVC indicators. 
Examples include \citet{dahuetal01}'s Vertical Specialisation ratio at the industry-level which can be derived
by summing for each country and industry across the value added of all foreign countries and industries.
Alternatively, the four dimensions of the matrix (source country, source industry, using country, using industry)
allow for industry-level gravity-type estimations of value added trade flows.

@@

\begin{Schunk}
\begin{Sinput}
> decompr_object <- load_tables_vectors( x = inter,
+                                        y = final,
+                                        k = countries,
+                                        i = industries,
+                                        o = out         )
\end{Sinput}
\end{Schunk}

We can now decompose the elements using the Leontief decomposition.
\begin{Schunk}
\begin{Sinput}
> lt <- leontief( decompr_object )
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> lt2 <- decomp( x = inter,
+                y = final,
+                k = countries,
+                i = industries,
+                o = out,
+                method = "leontief" )
\end{Sinput}
\end{Schunk}

@@

\subsection{Output}
\begin{Schunk}
\begin{Sinput}
> lt
\end{Sinput}
\begin{Soutput}
                              Argentina.Agriculture
Argentina.Agriculture                   28.52278143
Argentina.Textile.and.Leather            1.06206936
Argentina.Transport.Equipment            0.21043693
Turkey.Agriculture                       0.71952151
Turkey.Textile.and.Leather               0.41201175
Turkey.Transport.Equipment               0.03482652
Germany.Agriculture                      0.92530356
Germany.Textile.and.Leather              0.64666560
Germany.Transport.Equipment              0.66638333
                              Argentina.Textile.and.Leather
Argentina.Agriculture                            2.79395126
Argentina.Textile.and.Leather                   19.12053186
Argentina.Transport.Equipment                    0.14228369
Turkey.Agriculture                               1.34237213
Turkey.Textile.and.Leather                       1.38523849
Turkey.Transport.Equipment                       0.08553139
Germany.Agriculture                              2.25142713
Germany.Textile.and.Leather                      0.72785683
Germany.Transport.Equipment                      0.65080723
                              Argentina.Transport.Equipment Turkey.Agriculture
Argentina.Agriculture                            0.35606694         1.81066955
Argentina.Textile.and.Leather                    0.41813924         0.48370042
Argentina.Transport.Equipment                    1.06369578         0.03329456
Turkey.Agriculture                               0.11504126        34.92704803
Turkey.Textile.and.Leather                       0.11764036         2.69291816
Turkey.Transport.Equipment                       0.02667530         0.81210167
Germany.Agriculture                              0.16222512         2.31122022
Germany.Textile.and.Leather                      0.08244379         1.53837777
Germany.Transport.Equipment                      0.25807221         1.29066963
                              Turkey.Textile.and.Leather
Argentina.Agriculture                          3.1173841
Argentina.Textile.and.Leather                  1.8329024
Argentina.Transport.Equipment                  0.0790545
Turkey.Agriculture                             6.9994970
Turkey.Textile.and.Leather                    40.1671410
Turkey.Transport.Equipment                     0.9075189
Germany.Agriculture                            2.0595825
Germany.Textile.and.Leather                    2.5488967
Germany.Transport.Equipment                    1.4880228
                              Turkey.Transport.Equipment Germany.Agriculture
Argentina.Agriculture                         0.35901126           1.2364172
Argentina.Textile.and.Leather                 0.43058635           0.5937041
Argentina.Transport.Equipment                 0.04024626           0.0231846
Turkey.Agriculture                            1.47711579           2.5543089
Turkey.Textile.and.Leather                    1.31799873           1.1093993
Turkey.Transport.Equipment                    3.16041392           0.1151191
Germany.Agriculture                           0.51211484          29.8763359
Germany.Textile.and.Leather                   0.63316614           1.4593583
Germany.Transport.Equipment                   0.56934671           1.7321726
                              Germany.Textile.and.Leather
Argentina.Agriculture                          1.30283802
Argentina.Textile.and.Leather                  1.15375958
Argentina.Transport.Equipment                  0.07482343
Turkey.Agriculture                             1.52213499
Turkey.Textile.and.Leather                     1.15207241
Turkey.Transport.Equipment                     0.07448266
Germany.Agriculture                            5.24719728
Germany.Textile.and.Leather                   18.95868110
Germany.Transport.Equipment                    1.51401054
                              Germany.Transport.Equipment
Argentina.Agriculture                           4.1208736
Argentina.Textile.and.Leather                   4.7490351
Argentina.Transport.Equipment                   0.1932621
Turkey.Agriculture                              6.1806254
Turkey.Textile.and.Leather                      9.5069032
Turkey.Transport.Equipment                      0.6464733
Germany.Agriculture                             9.6006931
Germany.Textile.and.Leather                     8.1583150
Germany.Transport.Equipment                    34.7438192
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> w <- wwz(decompr_object)
\end{Sinput}
\end{Schunk}
